Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr

### 2018-01-24 GWAS-comparison.Rmd
save

```{r}
snpgdsClose(genofile)
# Need sudo yum install netcdf-devel.x86_64
library(GWASTools)
(genofile <- GdsGenotypeReader(gds, YchromCode=24L, XYchromCode=25L))
scanID <- getScanID(genofile)
# sex <- getVariable(genofile, "sample.annot/sex")
phenotype <- getVariable(genofile, "sample.annot/phenotype")
covar <- obj.svd$u
colnames(covar) <- paste0("PC", 1:10)
scanAnnot <- ScanAnnotationDataFrame(
  cbind.data.frame(scanID, pheno = phenotype - 1, covar)
)
snpID <- getSnpID(genofile)
chromosome <- getChromosome(genofile)
position <- getPosition(genofile)
alleleA <- getAlleleA(genofile)
alleleB <- getAlleleB(genofile)
snpAnnot <- SnpAnnotationDataFrame(data.frame(
  snpID = seq_along(snpID), chromosome, position, rsID = snpID,
  alleleA, alleleB, stringsAsFactors=FALSE),
  YchromCode=24L, XYchromCode=25L)

# Logistic regression
system.time(
  gwas3 <- assocRegression(genoData, "pheno", 
                           model.type = "logistic",
                           covar = colnames(covar),
                           snpEnd = 10e3)
) * ncol(G) / 10e3
all.equal(gwas3$Wald.pval, gwas.pval[rows_along(gwas3)]) 
# assocRegression uses glm() so that it is the most precise
# We are very precise and PLINK is the least precise
true.pval <- sapply(1:9, function(j) {
  summary(glm(y01 ~ G[, j] + covar, family = "binomial"))$coefficients[2, 4]
})
rbind(`true (glm)` = true.pval, bigstatsr = gwas.pval[1:9], 
      plink = gwas2$P[1:9], gwastools = gwas3$Wald.pval[1:9])
```

>>>>

```{r}
# PCA
system.time(
  pca <- snpgdsPCA(genofile, snp.id = ind.keep, eigen.cnt = 10,
                   num.thread = NCORES, algorithm = "randomized")
)
```

```{r}
# Format objects
snpgdsClose(genofile)
# Need sudo yum install netcdf-devel.x86_64
library(GWASTools)
(genofile <- GdsGenotypeReader(gds, YchromCode=24L, XYchromCode=25L))
scanID <- getScanID(genofile)
# sex <- getVariable(genofile, "sample.annot/sex")
phenotype <- getVariable(genofile, "sample.annot/phenotype")
covar <- obj.svd$u
colnames(covar) <- paste0("PC", 1:10)
scanAnnot <- ScanAnnotationDataFrame(
  cbind.data.frame(scanID, pheno = phenotype - 1, covar)
)
snpID <- getSnpID(genofile)
chromosome <- getChromosome(genofile)
position <- getPosition(genofile)
alleleA <- getAlleleA(genofile)
alleleB <- getAlleleB(genofile)
snpAnnot <- SnpAnnotationDataFrame(data.frame(
  snpID = seq_along(snpID), chromosome, position, rsID = snpID,
  alleleA, alleleB, stringsAsFactors=FALSE),
  YchromCode=24L, XYchromCode=25L)

genoData <- GenotypeData(genofile, scanAnnot = scanAnnot, snpAnnot = snpAnnot)
```

```{r}
# Logistic regression
system.time(
  gwas3 <- assocRegression(genoData, "pheno", 
                           model.type = "logistic",
                           covar = colnames(covar),
                           snpEnd = 10e3)
) * ncol(G) / 10e3
all.equal(gwas3$Wald.pval, gwas.pval[rows_along(gwas3)]) 
# assocRegression uses glm() so that it is the most precise
# We are very precise and PLINK is the least precise
true.pval <- sapply(1:9, function(j) {
  summary(glm(y01 ~ G[, j] + covar, family = "binomial"))$coefficients[2, 4]
})
rbind(`true (glm)` = true.pval, bigstatsr = gwas.pval[1:9], 
      plink = gwas2$P[1:9], gwastools = gwas3$Wald.pval[1:9])
```

### 2018-01-24 cabios-template2/bigsnpr-review.tex
save

\begin{table}[!tpb]
\begin{center}
\begin{adjustbox}{max width=235pt}
\begin{tabular}{|c|c|c|}
\hline
\multirow{2}{*}{Operation} &   \multicolumn{2}{c|}{Execution times (in seconds)} \\
 \cline{2-3}
 & PLINK 1.9 and FastPCA & bigstatsr and bigsnpr \\
\hline
Reading PLINK files & n/a & 5 / 20 \\
Pruning & 4 / 4 & 14 / 52 \\
Computing 10 PCs & 306 / 315 & 58 / 180 \\
GWAS (binary phenotype) & 339 / 293 & 301 / 861 \\
\hline
Total & 649 / 612 & 378 / 1113 \\
\hline
\end{tabular} 
\end{adjustbox}
\end{center}
\caption{Execution times with bigstatsr and bigsnpr compared to PLINK 1.9 and FastPCA (EIGENSOFT) for making a GWAS for the Celiac dataset. The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop (2 cores used and 8GB of RAM).} 
\label{tab:bench-gwas}
\end{table}

\begin{table}[!tpb]
\begin{center}
\begin{adjustbox}{max width=235pt}
\begin{tabular}{|c|c|c|}
\hline
\multirow{2}{*}{Operation} &   \multicolumn{2}{c|}{Execution times (in seconds)} \\
 \cline{2-3}
 & PRSice & bigstatsr and bigsnpr \\
\hline
Reading PLINK files &  & 6 / 20 \\
Reading summary stats &  & 4 /   \\
Clumping &  & 9 /  \\
PRS &  & 2 /  \\
Compute pvalues &  & 1 /  \\
\hline
Total & 22 /  & 22 /  \\
\hline
\end{tabular} 
\end{adjustbox}
\end{center}
\caption{Execution times with bigstatsr and bigsnpr compared to PRSice for making a PRS on the Celiac dataset based on summary statistics for Height. The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop (2 cores used and 8GB of RAM).}
\label{tab:bench-prs}
\end{table}

>>>>

\begin{table}[!tpb]
\begin{center}
\begin{adjustbox}{max width=235pt}
\begin{tabular}{|c|c|c|c|}
\hline
\multirow{3}{*}{Operation~~\textbackslash~~Software} &   \multicolumn{3}{c|}{Execution times (in seconds)} \\
 \cline{2-4}
 & FastPCA & bigstatsr  & SNPRelate \\
 & PLINK 1.9 & bigsnpr & GWASTools \\
\hline
Reading PLINK files & n/a & 6 / 20 & 12 /  \\
Pruning & 4 / 4 & 14 / 52 & 33 / \\  
Computing 10 PCs & 304 / 315 & 61 / 180 &  /  \\
GWAS (binary phenotype) & 336 / 293 & 301 / 861 & 15806 /  \\
GWAS (continuous phenotype) & 1339 /  & 9 /  & 5933 /  \\
\hline
Total & 649 / 612 & 378 / 1113 &  /  \\
\hline
\end{tabular} 
\end{adjustbox}
\end{center}
\caption{Execution times with bigstatsr and bigsnpr compared to PLINK 1.9 and FastPCA (EIGENSOFT) and also to R packages SNPRelate and GWASTools for making a GWAS for the Celiac dataset. The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop (2 cores used and 8GB of RAM).} 
\label{tab:bench-gwas}
\end{table}

\begin{table}[!tpb]
\begin{center}
\begin{adjustbox}{max width=235pt}
\begin{tabular}{|c|c|c|}
\hline
\multirow{2}{*}{Operation~~\textbackslash~~Software} &   \multicolumn{2}{c|}{Execution times (in seconds)} \\
 \cline{2-3}
 & PRSice & bigstatsr and bigsnpr \\
\hline
Reading PLINK files &  & 6 / 20 \\
Reading summary stats &  & 4 /   \\
Clumping &  & 9 /  \\
PRS &  & 2 /  \\
Compute pvalues &  & 1 /  \\
\hline
Total & 22 /  & 22 /  \\
\hline
\end{tabular} 
\end{adjustbox}
\end{center}
\caption{Execution times with bigstatsr and bigsnpr compared to PRSice for making a PRS on the Celiac dataset based on summary statistics for Height. The first execution time is with a desktop computer (6 cores used and 64GB of RAM) and the second one is with a laptop (2 cores used and 8GB of RAM).}
\label{tab:bench-prs}
\end{table}
